RED WINE by HANDE KESKIN SUNGUR
========================================================

I will discover the relationship between chemical properties and quality of 
wine using red wine data. The format includes Univarite, Bivariated and 
Multivariated analyses with a final summary and reflection at the end.

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(RColorBrewer)
library(gridExtra)
library(reshape)
library(GGally)
library(reshape2)
library(dplyr)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wineQualityReds <- read.csv('wineQualityReds.csv')
```


#Summary of the Data Set

```{r echo=FALSE, Summary_the_Data}
# Summary the Data
dim(wineQualityReds)
str(wineQualityReds)
summary(wineQualityReds)
```
- The red wine data set contains 1599 obersvations with 13 variables. 
11 of the variables are chemical features. 

- The range for fixed acidity is minimum 4.60 and maximum 15.90.

- The range for volatile acidity is minimum 0.12 and maximum 1.58.

- The range for volatile acidity is minimum 0 and maximum 1.

- The median quality is 6 and mean is 5.636. The quality of samples 
range 3 to 8.

- The median PH 3.310 and mean 3.311 . PH varies from 2.720 to 4.010 

# Univariate Plots Section

Examine histogram graphs of all values:

```{r echo=FALSE, Univariate_Plots}
p1<-ggplot(aes(x=fixed.acidity),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
    ggtitle('fixed acidity')
p2<-ggplot(aes(x=volatile.acidity),data =  wineQualityReds)+
   geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('volatile acidity')
p3<-ggplot(aes(x=citric.acid),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('citric acid')
p4<-ggplot(aes(x=residual.sugar),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('residual sugar')
p5<-ggplot(aes(x=chlorides),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('chlorides')
p6<-ggplot(aes(x=free.sulfur.dioxide),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('free sulfur dioxide')
p7<-ggplot(aes(x=total.sulfur.dioxide),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('total sulfur dioxide')
p8<-ggplot(aes(x=density),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('density')
p9<-ggplot(aes(x=pH),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('pH')
p10<-ggplot(aes(x=sulphates),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('sulphates')
p11<-ggplot(aes(x=alcohol),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('alcohol')
p12<-ggplot(aes(x=quality),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray")+
   ggtitle('quality')
grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12,ncol=3)

```

- PH value, density value , volatile acidity value exhibit quite 
normal distribution.
- Fixed acidity, citric acid, Free suldur dioxide, Total sulfur dioxide, 
Sulphates, Alchole value exhibit positively skewed distribution.
- Residual sugar, chlorides exhibit long tail distribution


### Quality of Red Wine 

The median quality is 6 and mean is 5.636. The quality of samples range 3 to 8.
Most of the quality ratings are either 5 or 6. The most of quality rating is 5.

```{r echo=FALSE, Quality_Count}
 table(wineQualityReds$quality)
 summary(wineQualityReds$quality)
```

Draw a histogram of the quality ratios: 

```{r echo=FALSE, Quality_Plots}
ggplot(aes(x=quality),data =  wineQualityReds)+
     geom_histogram(color =I('black'),fill = "dark gray",binwidth = 1)+
     ggtitle('quality distribution for Red wine')+ xlab("wine grade")+ 
                                                   ylab("count")


```

We creat a new variable that called "rating" which is categorically quality divided into "bad", "average", and "excellent".

** quality < 5 = 'bad'**
**quality < 7 = 'average'**
**quality > 6 = 'excellent'**

```{r echo=FALSE, Rating_Plots}
wineQualityReds$rating <- ifelse(wineQualityReds$quality < 5, 'bad', 
                  ifelse(wineQualityReds$quality < 7, 'average', 
                  ifelse(wineQualityReds$quality > 6, 'excellent', 'none')))

wqr<-wineQualityReds

wqr$rating <- ordered(wqr$rating, levels = c('bad', 'average', 'excellent'))

qplot(wqr$rating, xlab="quality")
```

### Level of Alcohol in Red Wine 

The median alcohol is 10.20 and mean is 10.42. 
The quality of samples range 8.40 to 14.90. 
Most of the quality ratings are either 5 or 6. The most of quality rating is 5.
Red Wine data sample is small but it gives the same pattern of alcohol level 
distribution as red wine.

```{r echo=FALSE, Alcohol_Count}
summary(wineQualityReds$alcohol)
ggplot(aes(x=alcohol),data =  wineQualityReds)+
     geom_histogram(color =I('black'),fill = "dark gray",binwidth = 1)+
     ggtitle('alcohol distribution for Red wine')+ xlab("level of alcohol")+ 
                                                   ylab("count")
```

### Level of Chlorides and Residual sugar in Red Wine 

Residual sugar, chlorides distribution is long tailed distribution. 
So I transformed this data for a more accurate distribution. 
The log10 produces a more understandable distribution for both.

```{r echo=FALSE, Chlorides_Residual_sugar_Plots}
grid.arrange(qplot(x = residual.sugar, data = wineQualityReds,binwidth = 0.1),
             qplot(x = log10(residual.sugar), data = wineQualityReds,binwidth=0.1),
             ncol = 1)

grid.arrange(qplot(x = chlorides, data = wineQualityReds,binwidth = 0.02),
             qplot(x = log10(chlorides), data = wineQualityReds,binwidth=0.02), 
             ncol = 1)
```

# Univariate Analysis

### What is the structure of your dataset?

There are 1599 red wine observation  with 13 variables in the dataset . 
11 of the variables are quantitative features .
this Features are PH, density, volatile acidity,
Fixed acidity ,citric acid, Free suldur dioxide, 
Total sulfur dioxide, Sulphates, Alchole.

The final variable of quantity scores the wine from 0 to 10. 
But Potential range from 3 to 8. 
All of the features have a minimum value greater than 0. 

### What is/are the main feature(s) of interest in your dataset?

The main features in the data set are quality. I would
like to determine which features are best for predicting the quality of a wine. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Alcohol,fixed acidity,residual sugar likely contribute to the quality of a wine. 

### Did you create any new variables from existing variables in the dataset?

I convert some of the continious variables into discrete range. I creat a new variable that called "rating" which is categorically divided into "low", "average", and "high". This grouping method will help me detect the difference among each group more easily.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The residual sugar histogram and Chlorides histogram did not look normal. 
I applied a log transform to x-axis.


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
#Drop unused column
wineQualityReds$X <- NULL

ggcorr(wineQualityReds, label = TRUE, 
       label_size = 3, label_round = 2, label_alpha = TRUE)   +ggtitle('correlation coefficients') 
```

The top **4** correlation coefficients with quality are:

**alchol-quality = 0.48**

**sulphates-quality = 0.26**

**citric.acid-quality = 0.22**

**fixed.acidity-quality = 0.12**

Alcohol content has a high correlation with red wine quality. 

The biggest negative corralation coefficients with quality are:

**volatile.acidity-quality = -0.39**

**total.sulfur.dioxide-quality = -0.19**

**density-quality = -0.17**

**chlorides-quality = -0.13**

Variables with the highest positive correlation include:

**fixed.acidity-citirc.acid = 0.67**

**fixed.acidity-density = 0.67**

**free.sulfur.dioxide-total.sulfur.dioxide = 0.67**

**alcohol-quality = 0.48**

**sulphates-chlorides = 0.37**

Variables with the highest positive correlation include:

**fixed.acidity-pH = -0.68**

**volatile.acidity-citirc.acid = -0.55**

**citric.acid-pH = -0.54**

**density-alcohol = -0.50**

**volatile.acidity-quality = -0.39**


```{r echo=FALSE, Bivariate_Plots_100}
c1 <- cor.test(wineQualityReds$fixed.acidity,wineQualityReds$quality,
               method='pearson')
c2 <- cor.test(wineQualityReds$volatile.acidity,wineQualityReds$quality,
               method='pearson')
c3 <- cor.test(wineQualityReds$citric.acid ,wineQualityReds$quality,
               method='pearson')
c4 <- cor.test(wineQualityReds$residual.sugar ,wineQualityReds$quality,
               method='pearson')
c5 <- cor.test(wineQualityReds$chlorides ,wineQualityReds$quality,
               method='pearson')
c6 <- cor.test(wineQualityReds$free.sulfur.dioxide,wineQualityReds$quality,
               method='pearson')
c7 <- cor.test(wineQualityReds$total.sulfur.dioxide ,wineQualityReds$quality,method='pearson')
c8 <- cor.test(wineQualityReds$density ,wineQualityReds$quality,
               method='pearson')
c9 <- cor.test(wineQualityReds$pH ,wineQualityReds$quality,
               method='pearson')
c10 <- cor.test(wineQualityReds$sulphates ,wineQualityReds$quality,
                method='pearson')
c11 <- cor.test(wineQualityReds$alcohol,wineQualityReds$quality,
                method='pearson')
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The biggest negative corralation coefficient with quality is volatile.acidity 
and The biggest positive corralation coefficient with quality is alcohol.
From the plot, quality increases at moderate rates with higher alcohol.
Red wine quality decreases as volatile acidity increases

```{r echo=FALSE, Bivariate_Plots_1}


ggplot(aes(x = rating, y = alcohol), data = wineQualityReds) +
  geom_boxplot() +
  stat_summary(fun.y = median, geom = 'point', shape = 4) +
  labs(x = 'quality') + ggtitle('Rating and Alcohol Relationship') 
by(wineQualityReds$alcohol, wineQualityReds$rating, summary) 
```

```{r echo=FALSE, Bivariate_Plots_2}
ggplot(aes(x = rating, y = volatile.acidity), data = wineQualityReds) +
   geom_boxplot() +
   stat_summary(fun.y = median, geom = 'point', shape = 4) +
   labs(x = 'quality') + ggtitle('Rating and  Volatile Acidity Relationship') 
by(wineQualityReds$volatile.acidity, wineQualityReds$rating, summary)
```

We plot pH and fixed acidity. The correlation coefficient is -0.67, 
meaning that pH tends to drop at fixed acidity increases, which makes sense.

```{r echo=FALSE, Bivariate_Plots_3}
ggplot(aes(x = fixed.acidity, y = pH), data = wineQualityReds) +
  geom_point(color = "orange", alpha = 1/4) + 
  stat_smooth(method = 'lm') + ggtitle('Ph and  Fixed acidity Relationship') 
cor(wineQualityReds$fixed.acidity, wineQualityReds$pH)
```

sulphate content is quite important for red wine quality, 
particularly for the highest quality levels including excellent quality .

```{r echo=FALSE,  Multivariate_Plots_50}
ggplot(wineQualityReds, aes(x=sulphates, fill=factor(rating))) + 
   geom_density(alpha=0.2) + ggtitle('Density and Sulphates Relationship') 

tapply(wineQualityReds$sulphates, factor(wineQualityReds$rating), summary)
```

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
I am surprised that the relationship between fixed.acidity and ph is the 
strongest relationship.

### What was the strongest relationship you found?

From the variables analyzed, the strongest relationship was between
fixed.acidity and pH, which had a correlation coefficient of -0.68.


# Multivariate Plots Section

I noticed that quality increased as sulphates increased, 
When comparing sulphates to alcohol.  
For excellent wines, alcohol played a important role in detecting 
quality given a sulphate level.

```{r echo=FALSE, Multivariate_Plots}

ggplot(data = wineQualityReds, aes(x = sulphates, y = alcohol, 
                  colour=factor( quality))) +   
  geom_point(  alpha=0.05) +

   geom_density2d(bins=2, size=2) + 
   scale_color_brewer() +   scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) + ggtitle('Alcohol vs  Sulphates') 

```

We can see higher quality wine have higher alcohol and lower volatile acidity.

```{r echo=FALSE, Multivariate_Plots_1}
redwine=wineQualityReds

redwine$alcohol.bucket = cut(redwine$alcohol,
                            c(8, 9, 10.5, 12, 16))

ggplot(aes(x = factor(quality), y = volatile.acidity), data = redwine) + 
   geom_boxplot(aes(fill = alcohol.bucket) )+ ggtitle('Volatile Acidity and Alcohol Bucket Relationship')
```

We can see higher quality wine have higher sulphates, higher citric acidity.

```{r echo=FALSE, Multivariate_Plots_2}
ggplot(aes(x = sulphates, 
           y = citric.acid  , color = factor(quality)), 
       data = redwine) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))+ ggtitle('Cidric Acid vs  sulphates')
```

There is no definitive evidence of the sugar 
content that is causing the bad wines.

```{r echo=FALSE, Multivariate_Plots_3}
ggplot(data = wineQualityReds,
       aes(x = residual.sugar, y = citric.acid,
           color = rating)) +
  coord_cartesian(xlim = c(0, 4)) +
  geom_point() +
  facet_wrap(~rating)+ ggtitle('Cidric Acid vs  Residual Sugar')
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

When looking at wine quality level, we see a positive 
relationship between alcohol and sulphates. also we see a 
negatif relation between quality and volatile.acidity.

### Were there any interesting or surprising interactions between features?

I am suprising that residual sugar has very little impact on wine quality.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x=quality),data =  wineQualityReds)+
    geom_histogram(color =I('black'),fill = "dark gray",binwidth = 1)+
   ggtitle('distribution of red wine quality') 


```

### Description One

This plot shows that the distribution of wine quality. 
You can show that dataset is unbalanced. 
It has many count for medium quality, 
but much fewer count on bad and excellent quality wine.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = volatile.acidity, 
           y = alcohol  , color = factor(quality)), 
       data = redwine) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))+
   ggtitle('Alcohol v Volatile Acidity') +theme_dark()

ggplot(aes(x = sulphates, 
           y = citric.acid  , color = factor(quality)), 
       data = redwine) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))+
   ggtitle('Citric Acid vs Sulphates') +theme_dark()


```

### Description Two
In general, high quality wine tend to have higher alcohol and 
lower volatile acidity content. They also tend to have 
higher sulphate and higher critic acid content.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = wineQualityReds,
       aes(x = log10(sulphates), y = alcohol,
           color = rating)) +
  geom_point(alpha = 0.6, size = 1.2, position = 'jitter') +
    facet_wrap(~rating) +
  labs(x = 'Sulphates', y = 'Alcohol') +
  ggtitle('Alcohol vs Sulphates') 

```

### Description Three

When sulphates were low, the wine was still rated bad. 
Low sulphate content appears to contribute to bad wines. 
Also average wines have higher concentrations of sulphates. 
Excellent wines have higher alcohol contentrations and 
higher sulphate contentrations. 

------

# Reflection

Red wine dataset contains information  on 1599 red wine that has got 
different chemical. initially I discover the relationship between 
chemical properties and quality of wine using red wine dataset. 
The wine quality is more complex. But plots and visuals make it 
easier to see where to explore more.

4 features that have the highest correlation coefficient 
with quality are alcohol, volatile acidity, sulphates,citric acid.
Alcohol content appeared to be the number one factor 
for determining an excellent wine. Additionally excellent 
red wine contains specific amount of Citric acid and sulfates. 
Volatile acidity has a negative correlation to wine quality and 
I am suprising that residual sugar has very little impact on wine quality.

First I understanding the individual variables in the data set, and then I 
explored different questions and leads as I continued to make observations on 
plots. I have successfully identified features that impact the quality of red
wine, visualized their relationships and summarize their statistics. 
I explored the quality of wines across many variables.Eventually I realised that
good wine is more than perfect combination of different chemical components.

There are very few wines that are rated as low or high quality. I could do a 
better analysis if I had more information about the wines of the upper and lower
classes. More information will certainly improve the accuracy of the prediction 
models. With this exploratory data analysis on the red wine dataset, I found the
biggest challenging was sharing the right amount of information. 
